Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-04-29 paper.tex
Update paper.tex
read up to line 1070 or so

\title{Software Architectures to Integrate Workflow Engines in Science Gateways}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We describe and evaluate 6
  architectures with metrics measuring integration challenge,
  robustness, extensibility, scalability and other specific
  features. Results show the pros and cons of the different
  architectures: tight integration and task encapsulation are the
  simplest architectures and are therefore the easiest to integrate
  and the most robust ones, most architectures perform the same in
  terms of extensibility, the pool model is the most scalable and
  meta-workflows are only available in nested workflows and workflow
  import. Our analysis provides insights for science gateway
  architects and workflow engine developers.
\end{abstract}

@@

Workflow engines are critical components in the ecosystem of tools and
services offered by science gateways for the efficient and transparent
exploitation of distributed infrastructures. Several
software architectures can be adopted to integrate workflow engines in
science gateways, with important consequences on the 
development effort required and on the resulting system. This paper describes,
exemplifies and compares such architectures, based on
system-independent representations of their main components and
interactions.

Our analysis is fed by experience gathered from the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years. It also builds on the lessons learned from several
science gateway and workflow projects, including
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
initiatives.

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework that highlight the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, we evaluate the architectures with
metrics measuring integration challenge, robustness, extensibility,
scalability and other specific features. The paper closes on a
discussion and conclusion that compare architectures and highlight
recommendations for science gateway architects and workflow engine developers.

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
a high throughput. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.Scalability is also a crucial concern
for such multi-user systems that manage substantial amounts of data
and execute computationally-intensive analyses. To this end, science
gateways may balance the load among different instances of the same
engine. They may also use advanced task scheduling policies on the
infrastructure, e.g. to ensure fairness among
users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

@@

Software interactions among these components are represented with
labeled, plain black arrows.  \emph{Abstract} interactions are
specific types of interactions that may be implemented using different
components and interactions. They are represented with dotted arrows
and labeled with \texttt{*}. An architecture that has an abstract
interaction is an abstract architecture that can be instantiated in
various ways.

Red color is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks created by the workflow
engine and executed on the infrastructure. Data represents any type of
file or database involved in workflow execution and stored on the
infrastructure. It does not cover workflow parameters such as strings,
numbers, etc. Administrator refers to a science gateway user with
extended privileges to install applications.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent}
workflow executed by the \emph{parent} engine is itself a workflow,
called \emph{child} workflow, that is executed by a \emph{child}
engine. The parent and child engines might use different workflow
description languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through abstract
interaction \texttt{*$_1$}. The science gateway also communicates with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions \texttt{*$_4$}
and \texttt{*$_5$}. The parent engine communicates with the child
engine through abstract interaction \texttt{*$_6$}. Administrator
installs workflows through interactions \texttt{a$_1$} and
\texttt{a$_2$}.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine..  \closednote{Rafael}{$S_2$
    would be 2 if the paradigm does not allow to distribute the
    orchestration? e.g., workflow executions implemented within MPI
    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
    it is very difficult to define ``does not allow to''. For
    instance, why couldn't you hack Pegasus PMC so much that it
    becomes possible to distribute workflow executions in several MPI
    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
    prefer to leave it like this, unless you have a suggestion to
    reflect this in the text.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.

>>>>

\title{Software Architectures for Integrating Workflow Engines in Science Gateways} \note{Naj}{'what' software architectures? what is their qualifier, adjective?}

@@

\begin{abstract}
  We investigate six software architectures commonly used to integrate
  workflow engines in science gateways. We describe and evaluate these software
  architectures with metrics that measure integration complexity,
  robustness, extensibility, scalability and other specific
  features\note{Naj}{Please be precise, what are they?}. Results illustrate the pros and cons of different
  architectures: Simple architectures with tight integration and task encapsulation are the
  the easiest to integrate and are most robust.Extensibility is equivalent in most architectures. The pool model is the most scalable one and meta-workflows are only available in nested workflows and workflow
  import. Our analysis provides insights for science gateway
  architects and workflow engine developers. \note{Naj}{I find this abstract to be vague, it can be made more concise and poignant}
\end{abstract}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools and
services offered by science gateways . Several
software architectures can be adopted to integrate workflow engines in
science gateways, however 
development effort required and the resulting system vary across different architectures. This paper and our analytical approach to comparison of software architectures is informed by our experince in the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years; as well as by lessons learned from several
science gateway and workflow projects, including
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and related
initiatives. This paper describes,
provides examples and compares such architectures, based on
system-independent representations of their main components and
interactions. In this section we provide background information and
definitions of different workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, different architectures are evaluated with
metrics measuring integration challenge, robustness, extensibility,
scalability and other specific features. We compare different architectures and highlight
recommendations for science gateway architects and workflow engine developers.

@@

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
a high throughput. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM), \note{Naj}{should we not mention PMP which is used for CIVET?}
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

In this paper, the working definition of a
\emph{workflow engine} (also abbreviated \emph{engine}) is "a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values." Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications. \note{provide examples?}

@@

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted. Scalability is also a crucial concern
for such multi-user systems that manage substantial amounts of data
and execute computationally-intensive analyses. To this end, science
gateways may balance the load among different instances of the same
engine. They may also use advanced task scheduling policies on the
infrastructure, e.g. to ensure fairness among
users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

@@

Software interactions among these components are represented with
labeled, plain black arrows.  \emph{Abstract} interactions are
specific types of interactions that may be implemented using different
components and interactions. \note{Naj}{three 'interactions' in one sentence} They are represented with dotted arrows
and labeled with \texttt{*}. An architecture that has an abstract
interaction is an abstract architecture that can be instantiated in
various ways.

Red color (WHERE?) is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks created by the workflow
engine and executed on the infrastructure. Data represents any type of
file or database involved in workflow execution and stored on the
infrastructure. It does not cover workflow parameters such as strings,
numbers, etc. Administrator refers to a science gateway user with
extended privileges to install applications. \note{Naj}{I propose a table of definitions.}

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent}
workflow executed by the \emph{parent} engine is itself a workflow,
called \emph{child} workflow, that is executed by a \emph{child}
engine. \note{Naj} {this reads funny} The parent and child engines might use different workflow
description languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through abstract
interaction \texttt{*$_1$}. The science gateway also communicates with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions \texttt{*$_4$}
and \texttt{*$_5$}. The parent engine communicates with the child
engine through abstract interaction \texttt{*$_6$}. Administrator
installs workflows through interactions \texttt{a$_1$} and
\texttt{a$_2$}.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \note{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort\note{Naj}{how do you metricize "nonesensical amount fo effors?"}. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine..  \closednote{Rafael}{$S_2$
    would be 2 if the paradigm does not allow to distribute the
    orchestration? e.g., workflow executions implemented within MPI
    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
    it is very difficult to define ``does not allow to''. For
    instance, why couldn't you hack Pegasus PMC so much that it
    becomes possible to distribute workflow executions in several MPI
    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
    prefer to leave it like this, unless you have a suggestion to
    reflect this in the text.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.
